---
title: "Class 8: Breast Cancer Mini Project"
author: "Anisa Mody (PID: A19145291)"
format: pdf
toc: TRUE
---

## Background
The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class.

Today, we will analyze a biopsy data set from a fine needle aspiration (FNA) of a breast mass that describe characteristics of the cell nuclei present in digitized images.

## Data Import

This data is made available as a CSV file for download. We can read this in using `read.csv()`:

```{r}
# Read in the Wisconsin breast cancer data from the CSV file
fna.data <- "WisconsinCancer.csv"

# Load the CSV file into R, using the first column as row names
wisc.df <- read.csv(fna.data, row.names=1)

# Preview the first 3 rows of the dataset
head(wisc.df, 3)
```


Make sure we remove or exclude the `diagnosis` column from the data set that we use for further analysis - this is the expert diagnosis as either M or B:

```{r}
# Remove the diagnosis column so PCA only uses numeric measurement variables
wisc.data <- wisc.df[,-1]

# Save the diagnosis labels as a separate factor variable
diagnosis <- as.factor(wisc.df$diagnosis)
```

## Exploratory Data Analysis

> Question 1. How many observations are in this dataset?

```{r}
# Count the number of observations in the dataset
nrow(wisc.data)
```

> Question 2. How many of the observations have a malignant diagosis?

```{r}
# Count how many samples are benign or malignant
table(wisc.df$diagnosis)
```

> Question 3. How many variables/features in this data are suffixed with `_mean`?

We can use the `grep()` function to help us here:

```{r}
# Show all column names in the numeric cancer dataset
colnames(wisc.data)
```

```{r}
# Count how many variables end with "_mean"
length(grep("_mean", colnames(wisc.data), value = T))
```

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale=TRUE` argument to `prcomp()`.

```{r}
# Run PCA on the numeric data and scale variables first
wisc.pr <- prcomp(wisc.data, scale=TRUE)

# Display a summary of variance explained by each principal component
summary(wisc.pr)
```

```{r}
# Loads ggplot2 for plotting
library(ggplot2)

#Plot PC1 against PC2
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point()
```

> Question 4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

```{r}
# Calculate the variance for each principal component
summary(wisc.pr)$importance[2, 1]
```


> Question 5.  How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
which(summary(wisc.pr)$importance[3, ] >= 0.70)[1]
```


> Question 6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
which(summary(wisc.pr)$importance[3, ] >= 0.90)[1]
```


> Question 7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

The plot is very cluttered, with many overlapping points and variable arrows, making it hard to read. It’s difficult to clearly distinguish patterns or individual observations because of the excessive labeling. While it contains useful information, the overall complexity makes it challenging to interpret quickly.

> Question 8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Plot PC1 against PC2
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

The PC1 vs PC2 plot typically shows a clearer separation between diagnosis groups, while the PC1 vs PC3 plot shows less distinct separation. This indicates that PC2 captures more of the variation relevant to distinguishing the groups than PC3 does.

**Variance**

```{r}
# Calculate the variance for each principal component
pr.var <- wisc.pr$sdev^2

#View the first few variance values
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Question 9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

```{r}
# Sort loadings by magnitude (largest first)
sort(abs(wisc.pr$rotation[,1]), decreasing = TRUE)[1:5]
```

```{r}
sort(abs(wisc.pr$rotation[,1]), decreasing = TRUE)
```

The loading for concave.points_mean on PC1 can be obtained with `wisc.pr$rotation["concave.points_mean", 1]`, and it has a relatively large value, indicating it strongly contributes to PC1. However, when comparing all loadings, there are a few features with even larger absolute values. This means that while it is important, it is not the single strongest contributor to PC1.

## Hierarchical Clustering

> Question 10. Using the `plot()` and `abline()` functions, what is the height at which the clustering model has 4 clusters?

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

# Calculate Euclidean distances between all samples
data.dist <- dist(data.scaled)

# Perform hierarchical clustering using complete linkage
wisc.hclust <- hclust(data.dist, method = "complete")

# Plot the clustering dendrogram
plot(wisc.hclust)
```



```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

# Calculate Euclidean distances between all samples
data.dist <- dist(data.scaled)

# Perform hierarchical clustering using complete linkage
wisc.hclust <- hclust(data.dist, method = "complete")

# Plot the clustering dendrogram
plot(wisc.hclust)

# Add a horizontal line around height 19, which gives about 4 clusters
abline(h = 19, col = "red", lty = 2)
```

> Question 12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
# Cluster using the first 7 principal components, which explain over 90% variance
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")

# Cut the PCA-based dendrogram into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)

# Compare PCA-based clusters to the expert diagnosis
table(wisc.pr.hclust.clusters, diagnosis)
```
I prefer ward.D2 because it minimizes within-cluster variance, which produces tighter and more compact clusters. In this dataset, it results in clearer separation between benign and malignant samples compared to complete linkage. Complete linkage usually forms more irregular clusters and shows weaker alignment with the known diagnoses.


## Combining Methods

```{r}
# Calculate the distance matrix using the first four principal components
d <- dist(wisc.pr$x[,1:4])

# Run hierarchical clustering on the PCA distance matrix using Ward's method
wisc.pr.hclust <- hclust(d, method = "ward.D2") 

# Plot the hierarchical clustering dendrogram
plot(wisc.pr.hclust)

#Add a horizontal line at height 70 to show where the tree can be cut into groups
abline(h=70, col="hotpink")
```

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```

How does this clustering `grps` correspond to the expert `diagnosis`?

```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```

```{r}
# Plot PC1 against PC2 with the points colored based on a vector `grps`
ggplot(wisc.pr$x) +
aes(PC1, PC2) +
geom_point(col=grps)
```

**Clustering on PCA Results**

> Question 13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
# Compare the original hierarchical clustering result to diagnosis
table(wisc.pr.hclust.clusters, diagnosis)
```

> Question 14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
# Create clusters from the scaled-data hierarchical clustering model
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

# Compare the non-PCA hierarchical clustering groups to the actual diagnoses
table(wisc.hclust.clusters, diagnosis)

# Compare the PCA-based clustering groups to the actual diagnoses again
table(grps, diagnosis)
```
The hierarchical clustering without PCA separates the diagnoses less effectively, with several clusters containing a mix of both benign and malignant samples. In contrast, the PCA-based clustering shows clearer grouping, with clusters more strongly dominated by one diagnosis type. This suggests that using PCA before clustering improves separation by reducing noise and focusing on the most informative variation in the data.

## Prediction

```{r}
# Save the URL for the new patient samples
url <- "https://tinyurl.com/new-samples-CSV"

# Read the new patient samples into R
new <- read.csv(url)

# Project the new samples into the existing PCA model
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
# Plot original samples using PC1 and PC2, colored by cluster group
plot(wisc.pr$x[,1:2], col=grps)

# Add the new samples as large blue points
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)

# Label the new samples as 1 and 2
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Question 16. Which of these new patients should we prioritize for follow up based on your results?

Sample 2 should be prioritized for follow-up because it lies within or closest to the cluster of malignant samples in the PCA plot. This suggests it shares similar characteristics with known cancerous cases. In contrast, sample 1 appears closer to the benign cluster and is therefore less concerning.
