---
title: "Class 5: Data Viz with ggplot2"
author: "Anisa Mody (PID: A19145291)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *base R graphics* like the `plot()` function and add on packages like **ggplot2**. 

Let's compare how we make a simple figure with these two systems:

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```

Before I can use ggplot2, I need to install it on my computer. To do this, we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our quarto doc (we only run it once in our R console) as it would re-install the package every time we render our quarto report.

Once installed, we need to load up the package into our R brain:

```{r}
library(ggplot2)
```


The main function in the **ggplot2** package is called `ggplot()`

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- First Layer: the **data** (a data.frame of the stuff we want to plot)
- Second Layer: the **aes**thetics (how the data maps to the plot)
- Third Layer: the **geom** layer (how you want the plot drawn, e.g. points, lines, bars, etc.)

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point()
```

## Add some custom features to the plot

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth() +
  theme_minimal() +
  labs(title = "Stopping Distance of Old Cars", x = "Speed (MPH)", y = "Distance (Feet)")

```

Q. Can you make the `geom_smooth()` function produce a linear line of best fit for the data and turn off the "gray" error region?

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Stopping Distance of Old Cars", x = "Speed (MPH)", y = "Distance (Feet)")
```

---
## Gene Expression Figure

Import the data to plot.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```

Initial Plot
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point()
```

Modified Plot
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point() +
  theme_bw() +
  labs(title = "Gene Expression Changes Upon GLP-1 Inhibitor Treatment", x = "Control (No Drug)", y = "Drug Treatment") + 
  scale_color_manual(values = c("hotpink", "gray", "slateblue1"))
```


## Going Further

Here we read the famous gapminder dataset:

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha=0.5) +
  theme_bw()
```


```{r}
p
```

I can add more layers to `p`, which is my base plot.

```{r}
p +
  facet_wrap(~gapminder$continent)
```


Make a plot for 1977 and 2007 only (not all the years in the dataset).

First use the **dplyr** package and the `filter()` function from that package to extract the year 2007.

```{r}
library(dplyr)
```


```{r}
gapminder_1977 <- filter(gapminder, year==1977)
ggplot(gapminder_1977) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha=0.5) +
  theme_bw() +
  labs(title = "Gapminder 1977")

gapminder_2007 <- filter(gapminder, year==2007)
ggplot(gapminder_2007) +
  aes(x = gdpPercap, y = lifeExp, col = continent) +
  geom_point(alpha=0.5) +
  theme_bw() +
  labs(title = "Gapminder 2007")
  
```

> Q. Make a histogram of lifeExp colored by continent

```{r}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) +
  geom_histogram(bins = 30) + 
  scale_fill_manual(values = c("violetred1", "salmon1", "navajowhite1", "lightgreen", "skyblue1"))
```


> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ continent)
  
  
```


