---
title: 'Class 12: Genome Informatics'
author: "Anisa Mody (PID: A19160915)"
output: pdf
toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1. Proportion of G/G in a Population 

Imported a CSV file from Ensembl < https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel > 

Here we will read in this CSV file:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

> Question 5 What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)? 

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

**14.0625%** of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G).

### GBR
> Now let's look at a different population, which is GBR (Great Britain), as stated in the lab sheet.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
gbr
```

> Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

Based on these results, it can be indicated that this variant, which is associated with childhood asthma, is more frequently found in the GBR population than in the MXL population. 

## Section 4: Population Scale Analysis [HOMEWORK]

> Question 13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

*Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding.*


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

**Total**

```{r}
nrow(expr)
```

**Sample sizes for each population**

```{r}
table(expr$geno)
```

**Median expression levels**

```{r}
colnames(expr) 
tapply(expr$exp, expr$geno, median)

```

These results show that there are 233 A|G genotypes, 108 A|A genotypes, and 121 G|G genotypes, which makes for a total of 462 genotypes. The median expression level is 31.248 for A|A, 25.064 for A|G, and 20.074 for G|G.

> Question 14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE) +
  scale_fill_manual(values = c(
    "A/A" = "lightpink",
    "A/G" = "lightyellow",
    "G/G" = "lightblue"
  )) +
  labs(
    title = "ORMDL3 Expression by Genotype",
    x = "Genotype",
    y = "Expression Level"
  ) +
  theme_minimal()
```

Analysis: The boxplot illustrates the distribution of ORMDL3 gene expression levels across the three genotypes: A/A, A/G, and G/G. The A/A genotype shows the highest median expression, A/G displays intermediate expression, and G/G has the lowest median expression, suggesting that the presence of the G allele is associated with progressively reduced ORMDL3 expression. Differences in spread and outliers among the groups indicate variation within each genotype, but the overall downward trend supports a potential functional effect of this SNP on gene regulation and expression levels.
